Towards practical dynamic induction control of wind farms: analysis of optimally controlled wind-farm boundary layers and sinusoidal induction control of first-row turbines

### 2018-02-14 draft.tex
Include Johan's comments

\begin{abstract}
	Wake interactions between wind turbines in wind farms lead to reduced energy extraction in downstream rows. In recent work, optimization and large-eddy simulation were combined with optimal dynamic induction control of wind farms to study the mitigation of these effects, showing potential power gains up to 20\% (Munters \& Meyers 2017 \emph{Phil Trans R Soc A} \textbf{375}, 20160100, doi:10.1098/rsta.2016.0100). However, the computational cost associated with the optimal control simulations impedes practical implementation of this approach. Furthermore, the resulting control signals optimally react to the specific instantaneous turbulent flow realizations in the simulations, so that they cannot be simply be used in general. The current work focuses on the detailed analysis of the optimization results of Munters \& Meyers, with the aim to identify simplified control strategies that mimic the optimal control results and can be used in practice. The analysis shows that wind-farm controls are optimized in a parabolic manner with little upstream propagation of information. Moreover, turbines can be classified into first-row, intermediate-row, and last-row turbines based on their optimal control dynamics. At the moment, the control mechanisms for intermediate-row turbines remain unclear, but for first-row turbines we find that the optimal controls increase wake mixing by periodic shedding of vortex rings. This behavior can be mimicked with a simple sinusoidal thrust control strategy for first-row turbines, resulting in robust power gains for turbines in the entrance region of the farm.
\end{abstract}

@@

In contrast to the studies cited above, that all focus on static setpoint optimization of wind farms, \cite{goit2015optimal} introduced a dynamic induction control approach based on large-eddy simulations (LES) and adjoint gradient optimization. In this study, individual turbines were used as dynamic flow actuators that influence the wind-farm boundary layer flow in such a way as to optimize aggregate wind-farm power extraction. The methodology was applied to the asymptotic case of a fully-developed `infinite' aligned wind farm, and power gains of about 16\% were quantified. Later, this approach was also used in induction control studies of wind farms with entrance effects in \cite{goit2016optimal} and, more recently, in \cite{munters2017optimal}, where similar gains in the order of 15\%--20\% were achieved. 
It is important to note that the computational cost of this LES-based dynamic induction control methodology is orders of magnitude too high for direct implementation. However, the methodology allows to assess the theoretical potential for wind-farm control, and an increased understanding of the flow physics of controlled wind farms can lead to simplified control strategies that can be applied in practice.

@@

A schematic overview of the wind-farm control methodology is shown in Figure \ref{fig:optimization_meth}. Figure \ref{fig:optimization_meth}a illustrates the control block diagram: an iterative optimization loop updates the wind-farm control vector $\bs{\varphi}(t)$ until a set of optimized controls $\bs{\varphi}^{\bullet}(t)$ is found. This optimization is based upon an unsteady turbulence-resolving LES wind-farm flow model, and sensitivities of the cost functional $\J$ (i.e. the total wind-farm power extraction) are calculated using an adjoint formulation of this flow model. In this way, a priori simplifications to the turbulent boundary layer and wake representation are avoided as much as possible, and the control signals are designed in a such a way that turbines actively tap into the dynamic of the turbulent flow. The optimization is performed using a receding-horizon control framework, as illustrated in Figure \ref{fig:optimization_meth}b. In this framework, wind-farm controls $\bs{\varphi}(t)$ are optimized for a finite time horizon $T$, involving a sequential set of LES and adjoint LES simulations. Upon convergence of the optimization, optimized control signals are applied in a flow advancement simulation for a time $T_A < T$, after which a new optimization window is initiated.

@@

In the remainder of this section, time-averaged wind-farm flow properties will be investigated. Here and throughout the remainder of this paper, we focus on case C3t5, as it produces similar energy gains as the highest-yield case C3t0 (see Figure \ref{fig:power_MM17}a), but achieves this with smooth thrust coefficient signals. In the following discussion, the time averaging operation is denoted by an overline, and flow field variables are decomposed into mean and fluctuating components as $\widetilde{u} = \overline{\widetilde{u}} + \widetilde{u}' \equiv \widetilde{U} + \widetilde{u}'  $. Figure~\ref{fig:flowfield_statistics} illustrates time-averaged flow field quantities of the reference case (left panels, a1 -- g1) and the differences between the optimized C3t5 case and the reference case (right panels, a2 -- g2). Simulation results are averaged over the different columns and are shown as either topviews at hub height (Figures \ref{fig:flowfield_statistics}b,f,g) or sideviews through a turbine column (Figures \ref{fig:flowfield_statistics}a,c,d,e).

@@

In order to further study how the optimal controls increase overall wind-farm power, Figure~\ref{fig:multirow} shows power extraction resulting from applying a subset of the optimal controls to specific turbine rows only. Figure~\ref{fig:multirow}a depicts simulation results for which the  optimized controls are applied \emph{only to one specific row}, with the thrust coefficient in all other rows kept at the reference value of $\cthat = 2$. From the figure it can be seen that only for the controls of the first row (R1) this results in a significant power increase in rows 2 and 3. This indicates that the optimal controls, as generated by the optimization at the wind-farm level, react to the precise flow conditions caused by upstream control actions, and can hence only be applied independently for the first row, which has no upstream dependence by definition.

@@

\subsubsection{Parameter sweep}
Instead of optimizing $A$ and $St$ using a similar gradient-based optimization setup as in MM17, we perform a parameter sweep to find optimal parameter combinations. The reason for this is twofold. On the one hand we would need a rather long optimization horizon $T$ to find a robust parameter combination that is independent of specific flow realizations. However, the chaotic nature of turbulent flow fields makes long-time optimization using adjoint LES practically infeasible to date (see, e.g. \citealp{wang2014least}). On the other hand the fact that we have only two parameters renders a parameter sweep computationally feasible. The sweep is performed for a reduced-size wind-farm LES, as illustrated in Figure~\ref{fig:sinus_setup}. The farm consists of $4 \times 4$ turbines in an aligned layout with $S = 6D$ in both streamwise and spanwise directions, geometrically equivalent to the optimally controlled wind farm in MM17. The simulation is performed on a domain of $4 \times 2.4 \times 1$ km$^3$ with a simulation grid of $192 \times 256 \times 144$ gridpoints. A wall roughness length $z_0 = 10^{-1}$ m is used. A set of wind-farm flow simulations is advanced in time by 30 minutes, during which the front row is controlled using a sinusoidally varying thrust coefficient $\cthat$, as defined in Equation \eqref{eq:define_thrust}. Within this set, the amplitude $A$ is varied between 0.5 and 2, with increments of 0.5. Furthermore, the Strouhal number $St$ is varied between 0.05 and 0.6, with increments of 0.05. In total, this leads to 48 LES cases within the set.

@@

Figure~\ref{fig:sinus_baseline} illustrates the power extraction for all cases considered. Figure~\ref{fig:sinus_baseline}a illustrates the relative power gains over the reference case. From the figure it can be seen that there is a well-defined range of values for $A$ and $St$ for which wind-farm power can be increased substantially through upstream sinusoidal thrust variations, with a maximal power increase of $\approx$ 5\% at $(St^\star, A^\star) = (0.25, 1.5)$. For instance, a Strouhal number $St = 0.25$ corresponds here to a sine wave period of $\approx$ 50 s for a turbine with diameter $D = 100$ m and a freestream velocity $U_\infty = 8.5$ m s$^{-1}$. The resulting maximum thrust coefficient of 3.5 can, e.g., be attained by an NREL 5MW rotor with a 50\% increase in chord length (see \citealp{goit2015optimal}, Appendix A). Figure~\ref{fig:sinus_baseline}b illustrates normalized power extraction by row for the reference case, the best sinusoidal case, and the first four rows of the optimal control case C3t5 from MM17. The figure shows that the power gain in the sinusoidal cases originates mostly from the second row, and that power in the first row is decreased by approximately 5\%. In contrast, optimal control case C3t5, in which all rows are active, also increases power in rows 3 and 4, and reduces first-row power by only 1\%. 
\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{gains_turbulent_6D_6D_6D_wide_inphase2.eps}
	\caption{Power extraction of baseline sinusoidal thrust case ($S = 6D$, $z_0 = 10^{-1}$ m). \emph{a) } Relative gain in mean wind-farm power extraction over reference case as a function of sine amplitude $A$ and frequency $St$. \emph{b) } Row-averaged mean power extraction for the best sinusoidal case, the optimal control case C3t5 from MM17, and the reference case, normalized by first-row reference power.\label{fig:sinus_baseline} }
\end{figure}

@@

\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{gains_fullscale.eps}
	\caption{Power extraction of full-scale sinusoidal thrust case ($S = 6D$, $z_0 = 10^{-1}$ m). \emph{a) } Relative gain in mean wind-farm power extraction over reference case as a function of sine amplitude $A$ and frequency $St$. \emph{b) } Row-averaged mean power extraction for the best sinusoidal case, the optimal control case C3t5 from MM17, and the reference case, normalized by first-row reference power.\label{fig:sinus_fullscale} }
\end{figure}

@@

As shown throughout the current section, a qualitative analysis of instantaneous flow features in the optimal induction control wind farm from MM17 has led to the identification of a sinusoidal thrust control strategy for first-row turbines, resulting in increased power extraction in the second row. In contrast to earlier work on steady axial induction control, for which power gains have often been marginal or non-existent (see, e.g. \citealp{annoni2016analysis}), the current analysis resulted in a robust axial induction control strategy with significant power gains, independent of specific local flow conditions. However, important comments should be made. First, sustained sinusoidal thrust variations with a large amplitude could contribute significantly to turbine fatigue loading. Hence, structural aspects should be taken into account upon evaluating the practical viability of the approach. Second, even though experiments have shown that, for practically relevant tip speed ratios, wind turbines shed vortices in a similar way as disk-like bluff bodies \citep{medici2006measurements}, the current behavior could still be an artifact of the relatively simple ADM used throughout this study. Further verification using higher fidelity wind turbine models, such as actuator line models, and wind-tunnel testing is hence necessary.

@@

First, as shown in the top panels of Figure~\ref{fig:cross_section_sinus}, even in the uncontrolled case, the kinetic energy of the flow in the vicinity of the turbine rotor is depleted more and more in the downstream rows. This complicates control strategies for these rows as the opportunity for increased mixing with high energy flow is decreased. Furthermore, intermediate turbines are subjected to increased turbulence levels and more complex vorticity dynamics, as illustrated in Figure~\ref{fig:vorticity_windfarm}. This could explain why the sinusoidal thrust control for the first row did not lead to increase power when applied to the second row: whereas the first row produces increased mixing by destabilizing relatively stable vortex sheets into vortex rings, the second row is already continuously immersed in complex vorticity patterns for which this simple approach does not work. 
However, note that for instance R7 in Figure~\ref{fig:controls} also seems to show quasi-periodic sinusoidal variations in $\cthat$ at a time period of approximately 50~s. This could be an indication that, also for intermediate rows, vortex ring shedding could amount to part of the power increase observed in the optimal control simulations, albeit at specific moments in time, synchronized with the local flow conditions.

Second, it is important to note that the vortex ring shedding mechanism constitutes only part of the power increase caused by the first row. Figure~\ref{fig:multirow} illustrates that the first-row optimized thrust coefficient also results in a significant power increase in the third row, which is not observed using the sinusoidal thrust strategy. Furthermore, the analysis of the modified control cases in Figure~\ref{fig:scrambled} proves that also the first-row controls are partially synchronized with the flow. This shows that other mechanisms, dependent on specific flow events for increasing wind-farm power, are at play as well. Even though the application of regression algorithms in an attempt to link turbine actions to low-dimensional flow measurements (e.g. local velocity, shear and kinetic energy) has been unsuccessful thus far, similar analysis based upon more complex flow features (e.g. vorticity structures, high-speed turbulent streaks, or downdrafts) might be more promising. However, this requires further optimal control simulations over an extended time, as the total control time horizon of 30 minutes in the current dataset is insufficient for robust statistics in this kind of analysis. This is an important remaining challenge to be addressed in future research.

>>>>

\begin{abstract}
	Wake interactions between wind turbines in wind farms lead to reduced energy extraction in downstream rows. In recent work, optimization and large-eddy simulation were combined with optimal dynamic induction control of wind farms to study the mitigation of these effects, showing potential power gains of up to 20\% (Munters \& Meyers 2017 \emph{Phil Trans R Soc A} \textbf{375}, 20160100, doi:10.1098/rsta.2016.0100). However, the computational cost associated with these optimal control simulations impedes practical implementation of this approach. Furthermore, the resulting control signals optimally react to the specific instantaneous turbulent flow realizations in the simulations, so that they cannot be simply used in general. The current work focuses on the detailed analysis of the optimization results of Munters \& Meyers, with the aim to identify simplified control strategies that mimic the optimal control results and can be used in practice. The analysis shows that wind-farm controls are optimized in a parabolic manner with little upstream propagation of information. Moreover, turbines can be classified into first-row, intermediate-row, and last-row turbines based on their optimal control dynamics. At the moment, the control mechanisms for intermediate-row turbines remain unclear, but for first-row turbines we find that the optimal controls increase wake mixing by periodic shedding of vortex rings. This behavior can be mimicked with a simple sinusoidal thrust control strategy for first-row turbines, resulting in robust power gains for turbines in the entrance region of the farm.
\end{abstract}

@@

In contrast to the studies cited above, that all focus on static setpoint optimization of wind farms, \cite{goit2015optimal} introduced a dynamic induction control approach based on large-eddy simulations (LES) and adjoint gradient optimization. In this study, individual turbines were used as dynamic flow actuators that influence the wind-farm boundary layer flow in such a way as to optimize aggregate wind-farm power extraction. The methodology was applied to the asymptotic case of a fully-developed `infinite' aligned wind farm, and power gains of about 16\% were quantified. Later, this approach was also used in induction control studies of wind farms with entrance effects in \cite{goit2016optimal} and, more recently, in \cite{munters2017optimal}, where similar gains in the order of 15\%--20\% were achieved. 
It is important to note that the computational cost of this LES-based dynamic induction control methodology is orders of magnitude too high for direct implementation as a practical control strategy. However, the methodology allows to assess the theoretical potential for wind-farm control, and increased understanding of the flow physics can lead to simplified control strategies that can be applied in practice.

@@

A schematic overview of the wind-farm control methodology is shown in Figure \ref{fig:optimization_meth}. Figure \ref{fig:optimization_meth}a illustrates the control block diagram: an iterative optimization loop updates the wind-farm control vector $\bs{\varphi}(t)$ until a set of optimized controls $\bs{\varphi}^{\bullet}(t)$ is found. This optimization is based upon an unsteady turbulence-resolving LES wind-farm flow model, and sensitivities of the cost functional $\J$ (i.e. the total wind-farm power extraction) are calculated using an adjoint formulation of this flow model. In this way, a priori simplifications to the turbulent boundary layer and wake representation are avoided as much as possible, and the control signals are designed in a such a way that turbines actively tap into the dynamics of the turbulent flow. The optimization is performed using a receding-horizon control framework, as illustrated in Figure \ref{fig:optimization_meth}b. In this framework, wind-farm controls $\bs{\varphi}(t)$ are optimized for a finite time horizon $T$, involving a sequential set of LES and adjoint LES simulations. Upon convergence of the optimization, optimized control signals are applied in a flow advancement simulation for a time $T_A < T$, after which a new optimization window is initiated.

@@

In the remainder of this section, time-averaged wind-farm flow properties will be investigated. Here and throughout the remainder of this paper, we focus on case C3t5, as it produces similar energy gains as the highest-yield case C3t0 (see Figure \ref{fig:power_MM17}a), but achieves this with smoother thrust coefficient signals. In the following discussion, the time averaging operation is denoted by an overline, and flow field variables are decomposed into mean and fluctuating components as $\widetilde{u} = \overline{\widetilde{u}} + \widetilde{u}' \equiv \widetilde{U} + \widetilde{u}'  $. Figure~\ref{fig:flowfield_statistics} illustrates time-averaged flow field quantities of the reference case (left panels, a1 -- g1) and the differences between the optimized C3t5 case and the reference case (right panels, a2 -- g2). Simulation results are averaged over the different columns and are shown as either topviews at hub height (Figures \ref{fig:flowfield_statistics}b,f,g) or sideviews through a turbine column (Figures \ref{fig:flowfield_statistics}a,c,d,e).

@@

In order to further study how the optimal controls increase overall wind-farm power, Figure~\ref{fig:multirow} shows power extraction resulting from applying a subset of the optimal controls to specific turbine rows only. Figure~\ref{fig:multirow}a depicts simulation results for which the  optimized controls are applied \emph{only to one specific row}, with the thrust coefficient in all other rows kept at the reference value of $\cthat = 2$. From the figure it can be seen that only for the controls of the first row (R1) this results in a significant power increase in rows 2 and 3. This indicates that the optimal controls, as generated by the optimization at the wind-farm level, react to the precise flow conditions caused by upstream control actions, and can hence only be applied independently for the first row, which has no upstream dependence on other controls.

@@

\subsubsection{Parameter sweep}
Instead of optimizing $A$ and $St$ using a similar gradient-based optimization setup as in MM17, we perform a parameter sweep to find optimal parameter combinations. The reason for this is that we would need a rather long optimization horizon $T$ to find a robust parameter combination that is independent of specific flow realizations. Unfortunately, the chaotic nature of turbulent flow fields makes long-time optimization using adjoint LES practically infeasible to date (see, e.g. \citealp{wang2014least}). However, the fact that we have only two parameters renders a parameter sweep computationally feasible. The sweep is performed for a reduced-size wind-farm LES, as illustrated in Figure~\ref{fig:sinus_setup}. The farm consists of $4 \times 4$ turbines in an aligned layout with $S = 6D$ in both streamwise and spanwise directions, geometrically equivalent to the optimally controlled wind farm in MM17. The simulation is performed on a domain of $4 \times 2.4 \times 1$ km$^3$ with a simulation grid of $192 \times 256 \times 144$ gridpoints. A wall roughness length $z_0 = 10^{-1}$ m is used. A set of wind-farm flow simulations is advanced in time by 30~minutes, during which the front row is controlled using a sinusoidally varying thrust coefficient $\cthat$, as defined in Equation \eqref{eq:define_thrust}. Within this set, the amplitude $A$ is varied between 0.5 and 2, with increments of 0.5. Furthermore, the Strouhal number $St$ is varied between 0.05 and 0.6, with increments of 0.05. In total, this leads to 48 LES cases within the set.

@@

Figure~\ref{fig:sinus_baseline} illustrates the power extraction for all cases considered. Figure~\ref{fig:sinus_baseline}a illustrates the relative power gains over the reference case. From the figure it can be seen that there is a well-defined range of values for $A$ and $St$ for which wind-farm power can be increased substantially through upstream sinusoidal thrust variations, with a maximal power increase of $\approx$ 5\% at $(St^\star, A^\star) = (0.25, 1.5)$. For instance, a Strouhal number $St = 0.25$ corresponds here to a sine wave period of $\approx$ 50 s for a turbine with diameter $D = 100$ m and a freestream velocity $U_\infty = 8.5$ m s$^{-1}$. The resulting maximum thrust coefficient of 3.5 can, e.g., be attained by an NREL 5MW rotor with a 50\% increase in chord length at a tip speed ratio that is 25\% higher than the greedy tip speed ratio (see \citealp{goit2015optimal}, Appendix A). Figure~\ref{fig:sinus_baseline}b illustrates normalized power extraction by row for the reference case, the best sinusoidal case, and the first four rows of the optimal control case C3t5 from MM17. The figure shows that the power gain in the sinusoidal cases originates mostly from the second row, and that power in the first row is decreased by approximately 5\%. In contrast, optimal control case C3t5, in which all rows are active, also increases power in rows 3 and 4, and reduces first-row power by only 1\%. 
\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{gains_turbulent_6D_6D_6D_wide_inphase2.eps}
	\caption{Power extraction of baseline sinusoidal thrust case ($S = 6D$, $z_0 = 10^{-1}$ m). \emph{a) } Relative gain in mean wind-farm power extraction over reference case as a function of sine amplitude $A$ and frequency $St$. \emph{b) } Row-averaged mean power extraction for the best sinusoidal case, the optimal control case C3t5 from MM17, and the reference case, normalized by first-row reference power.\label{fig:sinus_baseline} }
\end{figure}

@@

\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{gains_fullscale2.eps}
	\caption{Power extraction of full-scale sinusoidal thrust case ($S = 6D$, $z_0 = 10^{-1}$ m). \emph{a) } Relative gain in mean wind-farm power extraction over reference case as a function of sine amplitude $A$ and frequency $St$. \emph{b) } Row-averaged mean power extraction for the best sinusoidal case, the optimal control case C3t5 from MM17, and the reference case, normalized by first-row reference power.\label{fig:sinus_fullscale} }
\end{figure}

@@

As shown throughout the current section, a qualitative analysis of instantaneous flow features in the optimal induction control wind farm from MM17 has led to the identification of a sinusoidal thrust control strategy for first-row turbines, resulting in increased power extraction in the second row. However, important comments should be made. First, sustained sinusoidal thrust variations with a large amplitude could contribute significantly to turbine fatigue loading. Hence, structural aspects should be taken into account upon evaluating the practical viability of the approach. Second, even though experiments have shown that, for practically relevant tip speed ratios, wind turbines shed vortices in a similar way as disk-like bluff bodies \citep{medici2006measurements}, the current behavior could still be an artifact of the relatively simple ADM used throughout this study. Further verification using higher fidelity wind turbine models, such as actuator line models, and wind-tunnel testing is hence necessary.

@@

First, as shown in the top panels of Figure~\ref{fig:cross_section_sinus}, even in the uncontrolled case, the kinetic energy of the flow in the vicinity of the turbine rotor is depleted more and more in the downstream rows. This complicates control strategies for these rows as the opportunity for increased mixing with high energy flow is decreased. Furthermore, intermediate turbines are subjected to increased turbulence levels and more complex vorticity dynamics, as illustrated in Figure~\ref{fig:vorticity_windfarm}. This could explain why sinusoidal thrust control did not lead to increased power when applied to the second row: whereas the first row produces increased mixing by destabilizing relatively stable vortex sheets into vortex rings, the second row is already continuously immersed in complex vorticity patterns for which this simple approach does not work. 
However, note that for instance R7 in Figure~\ref{fig:controls} also seems to show quasi-periodic sinusoidal variations in $\cthat$ at a time period of approximately 50~s. This is an indication that, also for intermediate rows, vortex ring shedding could amount to part of the power increase observed in the optimal control simulations, albeit at specific moments in time, synchronized with the local flow conditions.

Second, it is important to note that the vortex ring shedding mechanism constitutes only part of the power increase caused by the first row. Figure~\ref{fig:multirow} illustrates that the first-row optimized thrust coefficient also results in a significant power increase in the third row, which is not observed using the sinusoidal thrust strategy. Furthermore, the analysis of the modified control cases in Figure~\ref{fig:scrambled} proves that also the first-row controls are partially synchronized with the flow. This shows that other mechanisms, dependent on specific flow events for increasing wind-farm power, are at play as well. Even though the application of regression algorithms in an attempt to link turbine actions to low-dimensional flow measurements (e.g. local velocity, shear and kinetic energy) has been unsuccessful thus far, similar analysis based upon more complex flow features (e.g. vorticity structures, high-speed turbulent streaks, or downdrafts) might be more promising. This requires further optimal control simulations over an extended time, as the total control time horizon of 30 minutes in the current dataset is insufficient for robust statistics in this kind of analysis. This is an important remaining challenge to be addressed in future research.
